% ROMJIST UniDive special issue - starter manuscript.
% Official class copied from template/Template_Latex/.../romjist.cls.
\documentclass[twoside]{romjist}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[top=4.3cm,bottom=5.7cm,left=3.74cm,right=3.75cm]{geometry}
\usepackage{times}
\usepackage{authblk}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{cite}
\usepackage[labelsep=space]{caption}

\title{Has Idiomaticity Representation Improved? A Controlled,
Model-Agnostic Re-Evaluation Across Model Generations}

\author[1]{M. Ali \uppercase{Bayram}}
\author[2]{Wei \uppercase{He}}
\author[3]{Aline \uppercase{Villavicencio}}
\author[ ]{\emph{[co-authors TBC after consultation]}}

\affil[1]{{\small Affiliation TBC}}
\affil[2]{{\small Affiliation TBC}}
\affil[3]{{\small Affiliation TBC}}
\affil[ ]{{\small \texttt{Email: TBC}}}

\begin{document}

% The following four values are for final ROMJIST print editing.
% Keep placeholders until acceptance/final page assignment.
\issuevolume{X}
\issuenumber{X}
\issueyear{2027}
\issuepages{XXX--XXX}

\headauthors{M. A. Bayram et al.}
\headtitle{Idiomaticity Representation Across Model Generations}

\maketitle

\begin{abstract}
He et al.\ (2025) showed that 2024-era word- and sentence-representation models do not
faithfully encode idiomaticity, relying instead on lexical overlap. We ask whether the rapid
progress of language models since then has changed this, and if not, what the underlying
obstacle is. We re-implement the original controlled evaluation---Noun Compound Idiomaticity
Minimal Pairs (NCIMP), the four probes, and the Affinity / Scaled-Similarity metrics---as a
model-agnostic framework, validate it against the published results, and apply it across three
cohorts: paper-era multilingual encoders, modern encoder and sentence-embedding models, and
an eleven-model Qwen/Gemma decoder cohort from 2025--2026. Across twenty model conditions, no
model reaches the partial-capture threshold. The best model, base Gemma 4-12B, reaches
ICS $=0.525$ and is only the first to prefer the gold idiomatic synonym over the literal
component. We also introduce the Ordinary-Calibrated Gap (OCG), which normalises idiomatic
synonym-vs-random separation by a model's behaviour on ordinary single-word and two-word
controls. The calibration shows that models separate ordinary synonyms from random
replacements much more cleanly than they separate idioms from literal alternatives. The results
support a structural-bottleneck account: scale narrows, but does not solve, idiomaticity
representation.
\end{abstract}

\begin{keywords}
Idiomaticity; multiword expressions; sentence embeddings; model evaluation; anisotropy; UniDive.
\end{keywords}

\section{Introduction}
% Motivation; the UniDive themes (multilingual/cross-lingual LT; quantifying diversity);
% research question (H1 scaling vs H0 structural bottleneck).

\section{Background and Related Work}
% He et al. (2025); NCIMP; probes; Affinity & Scaled Similarity; anisotropy.

\section{Method}
% Model-agnostic re-implementation; probes P_Syn/P_Comp/P_WordsSyn/P_Rand; metrics;
% improvement indicators ISC/IG/LOD/AID/FLOOR/RHO and composite ICS; cohorts; languages.

\section{The Ordinary-Calibrated Gap (OCG)}
% Definition: synonym-random gap of a group divided by the model's average synonym-random gap
% on ordinary single-word and two-word controls, at the same representation level
% (sentence / isolated-phrase / contextual-span). Purpose: separate idiomaticity-specific
% behaviour from generic lexical-overlap and anisotropy artefacts.

\section{Results}

\begin{table}[t]
\centering
\caption{English, paper-era vs modern encoder cohort (NC level).}
\begin{tabular}{lccc}
\toprule
Indicator & Paper-era & Modern & $\Delta$ \\
\midrule
ICS (composite $\uparrow$) & 0.341 & 0.430 & $+0.089\ (+26\%)$ \\
ISC ($\uparrow$)           & 0.099 & 0.174 & $+0.075$ \\
IG ($\downarrow$)          & 0.446 & 0.348 & $-0.098$ \\
LOD ($\downarrow$, $<0$)   & 0.227 & 0.077 & $-0.150$ \\
AID ($\uparrow$, $>0$)     & $-0.108$ & $-0.021$ & $+0.087$ \\
FLOOR (anisotropy)         & 0.575 & 0.723 & $+0.148$ \\
RHO ($\downarrow$)         & 0.615 & 0.578 & $-0.037$ \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]
\centering
\caption{Portuguese, paper-era vs modern encoder cohort (NC level).}
\begin{tabular}{lccc}
\toprule
Indicator & Paper-era & Modern & $\Delta$ \\
\midrule
ICS ($\uparrow$) & 0.428 & 0.423 & $-0.005\ (\approx 0\%)$ \\
ISC ($\uparrow$) & 0.150 & 0.095 & $-0.055$ \\
IG ($\downarrow$) & 0.304 & 0.328 & $+0.024$ \\
LOD ($\downarrow$) & 0.203 & 0.158 & $-0.045$ \\
AID ($\uparrow$) & $-0.085$ & $-0.026$ & $+0.059$ \\
FLOOR & 0.608 & 0.730 & $+0.122$ \\
RHO ($\downarrow$) & 0.421 & 0.492 & $+0.072$ \\
\bottomrule
\end{tabular}
\end{table}

% Add: per-model ICS; preliminary decoder-only LLM table; OCG cross-model figure.
% Source data: runs/en_all, runs/pt_all, article tables in
% article_perturbation_calibration/tables/.

\section{Discussion}
% Language-dependent, fragile gains; universal failure signatures; H0 supported.

\section{Conclusion and Future Work}
% Structural bottleneck; planned Turkish (morphologically rich) NCIMP extension; cross-architecture table.

\section*{Use of AI and AI-assisted technologies}
% Required disclosure statement per ROMJIST policy (language/readability assistance only).

\bibliographystyle{plain}
\bibliography{references}

\end{document}
